---
title: "16S rRNA Amplicon Analysis — Jeppe-Bac (MiSeq i100)"
author: "Justus Nweze"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    code_folding: show
    df_print: paged
    theme: flatly
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  fig.width = 8, fig.height = 5,
  cache = TRUE,          # expensive steps (learnErrors/dada/assignTaxonomy) cached
  cache.lazy = FALSE
)
```

# Overview

Denoising pipeline (DADA2 → phyloseq) for the **Jeppe-Bac** run: 26 paired-end
samples (JOS-Bac1..26), bacterial V3–V4 (Bac341F/805R), sequenced on a
**MiSeq i100**.

Because the i100 uses **XLEAP-SBS 2-colour chemistry** with **RTA4 binned
Q-scores**, three settings differ from a classic MiSeq pipeline:

1. **No `truncLen`** — quality stays high to the read end; use `minLen`.
2. **`rm.lowcomplex`** — 2-colour chemistry can miscall no-signal as high-Q
   "G", creating poly-G runs that distort the error model.
3. **Binned error model** — `makeBinnedQualErrfun()` instead of the default.

*(Source: `MiSeqi100_dada2_error_correction.pdf`.)*

```{r packages}
rm(list = ls())

setwd("C:/Users/jeppe/Desktop/Bioinf/MiSeq_8_7")
## Packages ##
library(BiocParallel)

## Force Bioconductor functions to run serially on Windows ##
BiocParallel::register(BiocParallel::SerialParam(), default = TRUE)
options(mc.cores = 1)

library(dada2)
library(phyloseq)
library(Biostrings)
library(ggplot2)
library(dplyr)

## Confirm ##
BiocParallel::bpparam()

```

# Paths and file names

```{r paths}

# the 52 .fastq.gz live here
path <- file.path("cutadapt_arch")
filt.dir <- file.path("filtered_arch")
out.dir  <- file.path("results_arch")
fig.dir  <- file.path("figures_arch")    # all plots exported here as PNG
dir.create(filt.dir, showWarnings = FALSE, recursive = TRUE)
dir.create(out.dir, showWarnings = FALSE, recursive = TRUE)
dir.create(fig.dir, showWarnings = FALSE, recursive = TRUE)

# Helper: display a ggplot AND save it to figures/ (300 dpi).
# Works both interactively (console) and when knitting.
save_fig <- function(p, name, w = 8, h = 5) {
  ggsave(file.path(fig.dir, name), plot = p, width = w, height = h,
         dpi = 300, bg = "white")
  p   # return so it also prints
}

fnFs <- sort(list.files(path, pattern = "_R1.fastq.gz", full.names = TRUE))
fnRs <- sort(list.files(path, pattern = "_R2.fastq.gz", full.names = TRUE))
# Sample name = Bac#_S###  (drop "JOS-" prefix, join the Bac token + S token)
# e.g. "JOS-Bac1_S146_L001_R1_001.fastq.gz" -> "Bac1_S146"  (matches metadata.csv)
parts <- strsplit(basename(fnFs), "_")
sample.names <- sapply(parts, function(x) paste0(sub("^JOS-", "", x[1]), "_", x[2]))
stopifnot(length(fnFs) == length(fnRs))
sample.names

length(fnFs)
length(fnRs)

basename(fnFs)
basename(fnRs)

all(file.exists(fnFs))
all(file.exists(fnRs))

file.info(fnFs)$size
file.info(fnRs)$size
```

# Sample display order and labels

The three sample sets are kept as separate objects so they cannot overwrite one
another. The matching set is selected automatically from the FASTQ sample names,
and the downstream plotting code continues to use `order` and `samlabels`.

```{r sample-order}
## Sample order ##
order_arch <- c(
  "Arch1_S136", "Arch2_S141", "Arch3_S145",
  "Arch13_S137", "Arch14_S138", "Arch15_S139",
  "Arch18_S140", "Arch20_S142", "Arch21_S143"
)

order_bac <- c(
  "Bac1_S146", "Bac2_S157", "Bac3_S165",
  "Bac4_S166", "Bac5_S167", "Bac6_S168",
  "Bac7_S169", "Bac8_S170", "Bac9_S171",
  "Bac10_S147", "Bac11_S148", "Bac12_S149",
  "Bac13_S150", "Bac14_S151", "Bac15_S152",
  "Bac16_S153", "Bac17_S154", "Bac18_S155",
  "Bac19_S156", "Bac20_S158", "Bac21_S159"
)

order_fh <- c(
  "JO-Bac1_S69", "JO-Bac2_S70", "JO-Bac3_S71",
  "JO-Bac4_S72", "JO-Bac5_S73", "JO-Bac6_S74",
  "JO-Bac7_S75", "JO-Bac8_S76"
)

## Plot labels ##
samlabels_arch <- c(
  "Arch1_S136"  = "BRt3_1 (oxic)",
  "Arch2_S141"  = "BRt3_2 (oxic)",
  "Arch3_S145"  = "BRt3_3 (oxic)",
  "Arch13_S137" = "UL_dir_1",
  "Arch14_S138" = "UL_dir_2",
  "Arch15_S139" = "UL_dir_3",
  "Arch18_S140" = "UL_SAt_3",
  "Arch20_S142" = "UL_SA_2",
  "Arch21_S143" = "UL_SA_3"
)

samlabels_bac <- c(
  "Bac1_S146"  = "BRt3_1 (oxic)",
  "Bac2_S157"  = "BRt3_2 (oxic)",
  "Bac3_S165"  = "BRt3_3 (oxic)",
  "Bac4_S166"  = "ULt2_1 (oxic)",
  "Bac5_S167"  = "ULt2_2 (oxic)",
  "Bac6_S168"  = "ULt2_3 (oxic)",
  "Bac7_S169"  = "BRt_1",
  "Bac8_S170"  = "BRt_2",
  "Bac9_S171"  = "BRt_3",
  "Bac10_S147" = "UL_1",
  "Bac11_S148" = "UL_2",
  "Bac12_S149" = "UL_3",
  "Bac13_S150" = "UL_dir_1",
  "Bac14_S151" = "UL_dir_2",
  "Bac15_S152" = "UL_dir_3",
  "Bac16_S153" = "UL_SAt_1",
  "Bac17_S154" = "UL_SAt_2",
  "Bac18_S155" = "UL_SAt_3",
  "Bac19_S156" = "UL_SA_1",
  "Bac20_S158" = "UL_SA_2",
  "Bac21_S159" = "UL_SA_3"
)

samlabels_fh <- c(
  "JO-Bac1_S69" = "FH1 (oxic)",
  "JO-Bac2_S70" = "FH2 (oxic)",
  "JO-Bac3_S71" = "FH3 (oxic)",
  "JO-Bac4_S72" = "FH no methane (oxic)",
  "JO-Bac5_S73" = "FH blank (oxic)",
  "JO-Bac6_S74" = "FH MeOH+NTA",
  "JO-Bac7_S75" = "FH MeOH+FeOH",
  "JO-Bac8_S76" = "FH MeOH only"
)

## Select the sample set that best matches the current FASTQ files ##
order_sets <- list(
  arch = order_arch,
  bac  = order_bac,
  fh   = order_fh
)

label_sets <- list(
  arch = samlabels_arch,
  bac  = samlabels_bac,
  fh   = samlabels_fh
)

match_counts <- vapply(
  order_sets,
  function(x) sum(sample.names %in% x),
  integer(1)
)

active_set <- names(which.max(match_counts))

if (max(match_counts) == 0) {
  stop("None of the predefined sample orders match the current FASTQ sample names.")
}

order <- order_sets[[active_set]]
samlabels <- label_sets[[active_set]]

## Guard against mismatched labels/order ##
stopifnot(all(order %in% names(samlabels)))

cat("Using sample order:", active_set, "\n")
```

# Read quality and complexity {.tabset}

MiSeq i100 profiles look flat/high (mostly Q38) and banded (binning) — that is
normal and is why we do not truncate.

## Forward quality
```{r qc-fwd}
save_fig(plotQualityProfile(fnFs[1:2]), "01_arch_quality_forward.png", w = 9, h = 5)
```

## Reverse quality
```{r qc-rev}
save_fig(plotQualityProfile(fnRs[1:2]), "02_arch_quality_reverse.png", w = 9, h = 5)
```

## Complexity (poly-G check)
Random sequence → ~16; poly-G → ~1. A peak near 1 means `rm.lowcomplex` (next
step) will strip those reads.
```{r qc-complex}
save_fig(plotComplexity(fnFs[1:2]), "03_arch_complexity_forward.png", w = 9, h = 5)
save_fig(plotComplexity(fnRs[1:2]), "04_arch_complexity_reverse.png", w = 9, h = 5)
```

# Filter and trim

Primers (still on reads): **Bac341F** `CCTACGGGNGGCWGCAG` (17 bp) /
**Bac805R** `GACTACHVGGGTATCTAATCC` (21 bp) → `trimLeft = c(17, 21)`.
MiSeq i100: `minLen` instead of `truncLen`, plus `rm.lowcomplex`.

```{r filter}
filtFs <- file.path(filt.dir, paste0(sample.names, "_F_filt.fastq.gz"))
filtRs <- file.path(filt.dir, paste0(sample.names, "_R_filt.fastq.gz"))
names(filtFs) <- sample.names
names(filtRs) <- sample.names

out <- filterAndTrim(
  fnFs, filtFs, fnRs, filtRs,
  minLen = 200,
  maxN = 0, maxEE = c(2, 2), truncQ = 2,
  rm.phix = TRUE, rm.lowcomplex = 8,
  compress = TRUE
)
out
```

> **JOS-Bac26 (S164)** is tiny (~59 KB) — almost certainly a negative control.
> Expect very few reads; keep it here to check contamination, drop before diversity.

# Error model (binned Q-scores) {.tabset}

```{r errors}
BinnedErrfun <- makeBinnedQualErrfun(c(2, 12, 24, 38))
errF <- learnErrors(filtFs, errorEstimationFunction = BinnedErrfun)
errR <- learnErrors(filtRs, errorEstimationFunction = BinnedErrfun)
```

## Forward errors
```{r err-fwd}
# flat line in the empty low-Q range is expected (no data in those bins)
save_fig(plotErrors(errF, nominalQ = TRUE), "05_arch_errors_forward.png", w = 8, h = 7)
```

## Reverse errors
```{r err-rev}
save_fig(plotErrors(errR, nominalQ = TRUE), "06_arch_errors_reverse.png", w = 8, h = 7)
```

# Denoise, merge, ASV table, chimeras

```{r denoise}
dadaFs <- dada(filtFs, err = errF)
dadaRs <- dada(filtRs, err = errR)

mergers <- mergePairs(dadaFs, filtFs, dadaRs, filtRs, verbose = TRUE)

seqtab <- makeSequenceTable(mergers)
table(nchar(getSequences(seqtab)))   # merged-length distribution (V3-V4 ~400-430)

seqtab.nochim <- removeBimeraDenovo(seqtab, method = "consensus", verbose = TRUE)
cat("Dimensions (samples x ASVs):", dim(seqtab.nochim), "\n")
cat("Fraction reads kept after chimera removal:",
    round(sum(seqtab.nochim) / sum(seqtab), 3), "\n")
```

# Read tracking (QC)

Big drops at any single step point to a parameter problem.

```{r track}
getN <- function(x) sum(getUniques(x))
track <- cbind(out,
  denoisedF = sapply(dadaFs, getN),
  denoisedR = sapply(dadaRs, getN),
  merged    = sapply(mergers, getN),
  nonchim   = rowSums(seqtab.nochim))
colnames(track)[1:2] <- c("input", "filtered")
rownames(track) <- sample.names
write.csv(track, file.path(out.dir, "read_tracking.csv"))
track
```

# Taxonomy

Taxonomy is assigned with the DADA2-formatted training FASTA specified below.
The downstream table keeps six taxonomic ranks through genus.

```{r taxonomy}

## GTDB r226 bacterial representative SSU training set ##
## Taxonomy is restricted to Domain -> Genus ##

gtdb.train <- "GTDB_r226_ar53_reps_DADA2_genus_trainset.fa.gz"

## Assign taxonomy ##
taxa.out <- assignTaxonomy(
  seqtab.nochim,
  gtdb.train,
  taxLevels = c(
    "Domain",
    "Phylum",
    "Class",
    "Order",
    "Family",
    "Genus"
  ),
  outputBootstraps = TRUE,
  multithread = 4,
  verbose = TRUE
)

## Taxonomy matrix ##
taxa <- taxa.out$tax

## Naive-Bayes bootstrap confidence (%) per rank ##
boot <- taxa.out$boot
colnames(boot) <- paste0(colnames(boot), " (BS)")

## Inspect taxonomy ##
taxa.print <- taxa
rownames(taxa.print) <- NULL

head(taxa.print)
```

# Save ASV outputs

```{r save-asv}
saveRDS(seqtab.nochim, file.path(out.dir, "seqtab_nochim_GTDB.rds"))
saveRDS(taxa,          file.path(out.dir, "taxa_GTDB.rds"))

asv_seqs <- colnames(seqtab.nochim)
asv_ids  <- paste0("ASV", seq_along(asv_seqs))

asv_tab <- t(seqtab.nochim)
rownames(asv_tab) <- asv_ids
write.csv(asv_tab, file.path(out.dir, "ASV_counts_arch_GTDB.csv"))

asv_tax <- cbind(taxa, boot)
rownames(asv_tax) <- asv_ids
write.csv(asv_tax, file.path(out.dir, "ASV_taxonomy_arch_GTDB.csv"))

fasta <- character(2 * length(asv_seqs))
fasta[c(TRUE, FALSE)] <- paste0(">", asv_ids)
fasta[c(FALSE, TRUE)] <- asv_seqs
writeLines(fasta, file.path(out.dir, "ASVs_GTDB.fasta"))
```

# phyloseq object

Build the phyloseq object used in the downstream analysis and attach the ASV
sequences. Sample names are kept identical to `sample.names`.

```{r phyloseq}
## Ensure sample names are correct ##
rownames(seqtab.nochim) <- sample.names

ps <- phyloseq(
  otu_table(seqtab.nochim, taxa_are_rows = FALSE),
  tax_table(taxa)
)

## Add ASV sequences ##
dna <- DNAStringSet(taxa_names(ps))
names(dna) <- taxa_names(ps)
ps <- merge_phyloseq(ps, dna)

## Rename taxa to ASV1, ASV2, ... ##
taxa_names(ps) <- paste0("ASV", seq_len(ntaxa(ps)))

saveRDS(ps, file.path(out.dir, "phyloseq_gtdb.rds"))
ps
```

# Community composition {.tabset}

```{r clean}
## Keep Bacteria/Archaea with assigned phylum ##
ps <- subset_taxa(
  ps,
  Domain %in% c("Bacteria", "Archaea")
)

ps <- subset_taxa(
  ps,
  !is.na(Phylum) & !Phylum %in% c("", "NA")
)

ps <- prune_taxa(taxa_sums(ps) > 0, ps)

sort(sample_sums(ps))
```


`rel_abund_bar()` agglomerates at the chosen rank and shows **every taxon that
reaches `min_pct`% in at least one sample as its own named bar** — so real genera
are never hidden inside a generic "Other". `Unassigned` (no genus in GTDB) is
kept as its own honest category. Only the genuinely negligible tail (each taxon
< `min_pct`%) collapses into a single grey `<min_pct%` block; each bar still sums
to 100%.

At `min_pct = 1` these 4 samples show ~13 genera. Note: past ~8 hues, colours are
no longer all colour-blind-distinct, so identity leans on the legend + the white
segment gaps. Lowering `min_pct` reveals more taxa but pushes colour reuse
further (there are 101 genus-level groups here, ~87 of them < 1%).

```{r relabund-fn}
# Distinct palette for up to 16 named taxa. The first 8 are the validated
# colour-blind-safe set; the rest extend it (less CVD-separated — lean on the
# legend). Grey (#c4c3bb) is reserved for the "<min_pct%" remainder.
tax_pal <- c(
  "#1F77B4", # blue
  "#FF7F0E", # orange
  "#2CA02C", # green
  "#D62728", # red
  "#9467BD", # purple
  "#8C564B", # brown
  "#E377C2", # pink
  "#7F7F7F", # gray
  "#BCBD22", # olive
  "#17BECF", # cyan
  "#393B79", # navy
  "#637939", # dark olive
  "#8C6D31", # ochre
  "#843C39", # dark red
  "#7B4173", # plum
  "#3182BD", # sky blue
  "#31A354", # emerald
  "#756BB1", # lavender
  "#636363", # charcoal
  "#E6550D", # burnt orange
  "#969696", # light gray
  "#6BAED6", # light blue
  "#74C476", # light green
  "#FD8D3C", # peach
  "#9ECAE1", # pale blue
  "#A1D99B", # pale green
  "#FDD0A2", # light orange
  "#C994C7", # lilac
  "#BDBDBD", # silver
  "#F768A1", # magenta
  "#41AB5D", # forest green
  "#08519C"  # deep blue
)


# Show every taxon reaching >= min_pct in ANY sample as its own bar; taxa below
# that (each individually tiny) collapse into one "<min_pct%" block. Unassigned
# (no genus in GTDB) is kept as its own category.
rel_abund_bar <- function(ps, level = "Genus", min_pct = 1) {
  rel <- transform_sample_counts(ps, function(x) x / sum(x))
  g   <- tax_glom(rel, taxrank = level, NArm = FALSE)
  df  <- psmelt(g)
  df[[level]] <- as.character(df[[level]])
  df[[level]][is.na(df[[level]])] <- "Unassigned"
  df$Sample <- factor(df$Sample, levels = order)
  mx   <- tapply(df$Abundance, df[[level]], max)              # peak abundance per taxon
  keep <- names(mx)[mx >= min_pct / 100]
  ord  <- names(sort(tapply(df$Abundance, df[[level]], mean), decreasing = TRUE))
  keep <- ord[ord %in% keep]                                 # kept taxa, abundance-ordered
  if (length(keep) > length(tax_pal))
    warning(length(keep), " taxa >= min_pct but only ", length(tax_pal),
            " palette colours; raise min_pct or extend tax_pal.")
  rare <- paste0("<", min_pct, "%")
  df$Taxon <- factor(ifelse(df[[level]] %in% keep, df[[level]], rare),
                     levels = c(keep, rare))
  df <- aggregate(Abundance ~ Sample + Taxon, df, sum)       # one block per taxon (no striping)
  fills <- c(setNames(tax_pal[seq_along(keep)], keep), setNames("#c4c3bb", rare))
  ggplot(df, aes(Sample, Abundance, fill = Taxon)) +
    geom_col(width = 0.95, colour = "white", linewidth = 0.15) +
    scale_fill_manual(values = fills) +
    scale_x_discrete(labels = samlabels) +
    scale_y_continuous(labels = function(x) x * 100, expand = c(0, 0)) +
    labs(x = NULL, y = "Relative abundance (%)", fill = level) +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
          legend.key.size = unit(0.42, "cm"), legend.text = element_text(size = 8))
}
```

### Phylum (overview)
```{r relabund-phylum, fig.height=6}
save_fig(rel_abund_bar(ps, level = "Phylum", min_pct = 1),
         "07_arch_gtdb_relabund_phylum.png", w = 7, h = 6)
```

### Genus (detail)
```{r relabund-genus, fig.height=7}
save_fig(rel_abund_bar(ps, level = "Genus", min_pct = 2),
         "08_arch_gtdb_relabund_genus.png", w = 9, h = 7)
```

# Methanotroph / ANME genus subset

The original script contained both an aerobic MOB list and an ANME list assigned
to the same object. To preserve the behaviour of the original code, the aerobic
MOB list is retained as a reference object, while `mob_genera` remains the ANME
list that was active after the overwrite.

The abundance plots below are therefore calculated from the active
`mob_genera` list as a fraction of the whole community.

```{r methanotroph-list}
aerobic_mob_genera <- c(
  ## Gammaproteobacteria — Methylococcales
  "Methylicorpusculum",
  "Methylococcus",
  "Methylocaldum",
  "Methylomonas",
  "Methylobacter",
  "Methylomicrobium",
  "Methylotuvimicrobium",
  "Methylosarcina",
  "Methylosoma",
  "Methylovulum",
  "Methyloglobulus",
  "Methylomarinum",
  "Methyloprofundus",
  "Methylotetracoccus",
  "Methylospira",
  "Methylomagnum",
  "Methyloterricola",
  "Methyloparacoccus",
  "Methylogaea",
  "Methylocucumis",
  "Methylohalobius",
  "Methylothermus",
  "Methylumidiphilus",
  "Crenothrix",
  "Clonothrix",
  "UBA4132",

  ## Alphaproteobacteria — Type II
  "Methylocystis",
  "Methylosinus",
  "Methylocella",
  "Methylocapsa",
  "Methyloferula",
  "Methylovirgula",
  "Methylorosula",

  ## Verrucomicrobia
  "Methylacidiphilum",
  "Methylacidimicrobium",

  ## NC10
  "Methylomirabilis"
)

mob_regex <- paste0(
  "^(",
  paste(aerobic_mob_genera, collapse = "|"),
  ")(_[A-Z0-9]+)?$"
)

```

```{r methanotroph-bar, fig.height=5}
# % of the WHOLE community: transform first, then keep MOB genera.
rel_all <- transform_sample_counts(ps, function(x) x / sum(x))
mob_df <- psmelt(rel_all)
mob_df$Genus <- as.character(mob_df$Genus)
mob_df <- mob_df[!is.na(mob_df$Genus) & grepl(mob_regex, mob_df$Genus), ]
mob_df <- mob_df[!mob_df$Sample %in% c("Bac22_S160", "Bac23_S161", "Bac24_S162", "Bac25_S163"), ]
mob_df$Sample <- factor(mob_df$Sample, levels = order)
  
 if (nrow(mob_df) == 0) {
  message("No methanotroph genera detected in these samples.")
} else {
  mob_df <- aggregate(Abundance ~ Sample + Genus, mob_df, sum)  # one block per genus
  ord <- names(sort(tapply(mob_df$Abundance, mob_df$Genus, mean), decreasing = TRUE))
  mob_df$Genus <- factor(mob_df$Genus, levels = ord)
  fills  <- setNames(tax_pal[seq_along(ord)], ord)

  p_mob <- ggplot(mob_df, aes(Sample, Abundance, fill = Genus)) +
    geom_col(width = 0.9, colour = "white", linewidth = 0.15) +
    scale_fill_manual(values = fills) +
    scale_y_continuous(labels = function(x) x * 100, expand = c(0, 0.01)) +
    scale_x_discrete(labels = samlabels) +
    labs(x = NULL, y = "Relative abundance (% of community)",
         fill = "ANME genus") +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
          legend.key.size = unit(0.42, "cm"), legend.text = element_text(size = 8))
  save_fig(p_mob, "09_arch_gtdb_relabund_methanotrophs.png", w = 7, h = 5)

  # Companion table: total methanotroph % per sample (written to results/).
  tot <- aggregate(Abundance ~ Sample, mob_df, sum)
  tot$Percent <- round(100 * tot$Abundance, 3)
  write.csv(tot[, c("Sample", "Percent")],
            file.path(out.dir, "methanotroph_total_percent.csv"), row.names = FALSE)
  tot[, c("Sample", "Percent")]
  
}

plot(p_mob)
```

## Within the selected methanotroph community

The same active genus set, but now normalised so the selected methanotroph
community sums to 100% within each sample.

```{r methanotroph-within, fig.height=5}
# Keep only MOB taxa, THEN transform so each sample's MOB sum to 1.
mob_taxa <- grepl(mob_regex, as.character(tax_table(ps)[, "Genus"]))
mob_taxa[is.na(mob_taxa)] <- FALSE

if (!any(mob_taxa)) {
  message("No methanotroph genera detected in these samples.")
} else {
  ps_mob <- prune_taxa(mob_taxa, ps)
  ps_mob <- transform_sample_counts(ps_mob, function(x) x / sum(x))   # sums to 1 per sample
  win_df <- psmelt(ps_mob)
  win_df <- win_df[!win_df$Sample %in% c("Bac22_S160", "Bac23_S161", "Bac24_S162", "Bac25_S163"), ]
  win_df$Sample <- factor(win_df$Sample, levels = order)
  win_df$Genus <- as.character(win_df$Genus)
  win_df <- aggregate(Abundance ~ Sample + Genus, win_df, sum)        # one block per genus
  ord <- names(sort(tapply(win_df$Abundance, win_df$Genus, mean), decreasing = TRUE))
  win_df$Genus <- factor(win_df$Genus, levels = ord)
  fills  <- setNames(tax_pal[seq_along(ord)], ord)

  p_win <- ggplot(win_df, aes(Sample, Abundance, fill = Genus)) +
    geom_col(width = 0.9, colour = "white", linewidth = 0.15) +
    scale_fill_manual(values = fills) +
    scale_y_continuous(labels = function(x) x * 100, expand = c(0, 0)) +
    scale_x_discrete(labels = samlabels) +
    labs(x = NULL, y = "Relative abundance (% of methanotrophs)",
         fill = "Methanotroph genus") +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
          legend.key.size = unit(0.42, "cm"), legend.text = element_text(size = 8))
  save_fig(p_win, "10_arch_gtdb_methanotrophs_within.png", w = 7, h = 5)

  # Companion table: each genus as % of the methanotroph community (wide).
  win_wide <- reshape(transform(win_df, Percent = round(100 * Abundance, 3))[,
                        c("Sample", "Genus", "Percent")],
                      idvar = "Sample", timevar = "Genus", direction = "wide")
  names(win_wide) <- sub("^Percent\\.", "", names(win_wide))
  write.csv(win_wide, file.path(out.dir, "methanotroph_within_percent.csv"),
            row.names = FALSE)
  win_wide
}
```

```{r}
## Named ANME genera ##
anme_genera <- c(
  "Methanophaga",
  "Candidatus Methanophaga",
  "Methanoalium",
  "Candidatus Methanoalium",
  "Methanoxibalbensis",
  "Candidatus Methanoxibalbensis",
  "Methanospirans",
  "Candidatus Methanospirans",
  "Methanospirare",
  "Candidatus Methanospirare",
  "Methanocomedens",
  "Candidatus Methanocomedens",
  "Methanomarinus",
  "Candidatus Methanomarinus",
  "Methanogaster",
  "Candidatus Methanogaster",
  "Methanoperedens",
  "Candidatus Methanoperedens",
  "Methanovorans",
  "Candidatus Methanovorans"
)

## Exact genus matching ##
anme_regex <- paste0(
  "^(",
  paste(anme_genera, collapse = "|"),
  ")(_[A-Z0-9]+)?$"
)

## Relative abundance of whole community ##
rel_all <- transform_sample_counts(
  ps,
  function(x) x / sum(x)
)

anme_df <- psmelt(rel_all)

anme_df$Genus <- as.character(anme_df$Genus)

anme_df <- anme_df[
  !is.na(anme_df$Genus) &
    grepl(anme_regex, anme_df$Genus),
]

## ---------------------------------------------------------
## If no ANME are detected, explicitly create zero values
## ---------------------------------------------------------

if (nrow(anme_df) == 0) {

  anme_plot_df <- data.frame(
    Sample = sample_names(ps),
    Abundance = 0
  )

  anme_plot_df$Sample <- factor(
    anme_plot_df$Sample,
    levels = order
  )

  p_anme <- ggplot(
    anme_plot_df,
    aes(x = Sample, y = Abundance)
  ) +
    geom_col(
      width = 0.9,
      fill = "grey70",
      colour = "white",
      linewidth = 0.15
    ) +
    scale_y_continuous(
      labels = function(x) x * 100,
      limits = c(0, 0.01),
      expand = c(0, 0)
    ) +
    scale_x_discrete(labels = samlabels) +
    labs(
      x = NULL,
      y = "ANME relative abundance (% of community)",
      title = "Anaerobic methanotrophic archaea",
      subtitle = "No ANME detected"
    ) +
    theme_bw() +
    theme(
      axis.text.x = element_text(
        angle = 90,
        hjust = 1,
        vjust = 0.5
      )
    )

} else {

  ## Normal plotting when ANME are present ##
  anme_df <- aggregate(
    Abundance ~ Sample + Genus,
    anme_df,
    sum
  )

  anme_df$Sample <- factor(
    anme_df$Sample,
    levels = order
  )

  ord <- names(
    sort(
      tapply(
        anme_df$Abundance,
        anme_df$Genus,
        mean
      ),
      decreasing = TRUE
    )
  )

  anme_df$Genus <- factor(
    anme_df$Genus,
    levels = ord
  )

  fills <- setNames(
    tax_pal[seq_along(ord)],
    ord
  )

  p_anme <- ggplot(
    anme_df,
    aes(
      Sample,
      Abundance,
      fill = Genus
    )
  ) +
    geom_col(
      width = 0.9,
      colour = "white",
      linewidth = 0.15
    ) +
    scale_fill_manual(values = fills) +
    scale_y_continuous(
      labels = function(x) x * 100,
      expand = c(0, 0.01)
    ) +
    scale_x_discrete(labels = samlabels) +
    labs(
      x = NULL,
      y = "ANME relative abundance (% of community)",
      fill = "ANME genus"
    ) +
    theme_bw() +
    theme(
      axis.text.x = element_text(
        angle = 90,
        hjust = 1,
        vjust = 0.5
      ),
      legend.key.size = unit(0.42, "cm"),
      legend.text = element_text(size = 8)
    )
}

save_fig(
  p_anme,
  "09_arch_relabund_ANME.png",
  w = 7,
  h = 5
)

p_anme
```

# Session info

```{r sessioninfo}
sessionInfo()
```
